---
title: "ESM 244 Lab 3 part 1: working with spatial data"
author: "Casey O'Hara"
format: 
  html:
    code-fold: show
    toc: true
    number-sections: true
    embed-resources: true
editor: visual
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
library(here)
library(broom)

### spatial packages
library(sf)
library(tmap)
```

## Read in our data

```{r}
ca_counties_raw_sf <- read_sf(here('data', 'ca_counties', 'CA_Counties_TIGER2016.shp'))

ca_counties_sf <- ca_counties_raw_sf %>%
  janitor::clean_names() %>%
  mutate(land_km2 = aland / 1e6) %>%
  select(county = name, land_km2)

ca_counties_df <- ca_counties_sf %>%
  as.data.frame() %>%
  select(-geometry)
```

## check the CRS

```{r}
ca_counties_sf %>% st_crs()
ca_counties_sf %>% terra::crs()
```

## explore it visually

```{r}
plot(ca_counties_sf %>% select(land_km2))
```

```{r}
ggplot() +
  geom_sf(data = ca_counties_sf, aes(fill = land_km2), color = 'white', size = 0.1) +
  theme_void() +
  scale_fill_gradientn(colors = c('cyan', 'blue', 'brown','purple'))
```

## Read in some sesbania records

```{r}
sesbania_sf <- read_sf(here('data/red_sesbania/ds80_for_lab.gpkg')) %>%
  janitor::clean_names()

sesbania_sf %>% st_crs()

plot(sesbania_sf %>% select(id))
```

## The analysis: 

Let's find the count of red sesbania observed locations in this dataset, by county, and then create a map of all CA counties using the fill color to indicate sesbania counts.

### Pseudocode:

full join!  
getting the two data sets into the same coordinate reference system
get county boundary coordinates to match up with the coordinates of sesbania
plot the datasets together
spatial join - matches up the datasets

## Transform the CRS so records match counties

```{r}
sesbania_3857_sf <- st_transform(sesbania_sf, 3857)
sesbania_3857_2_sf <- st_transform(sesbania_sf, st_crs(ca_counties_sf))

st_crs(sesbania_3857_sf)
```

## Let's plot the two together!

```{r}
ggplot() +
  geom_sf(data = ca_counties_sf) +
  geom_sf(data = sesbania_3857_sf, size = 1, color = 'red')
```

## Spatial join!

```{r}
ca_sesb_sf <- st_join(ca_counties_sf, sesbania_3857_sf)

sesb_ca_sf <- st_join(sesbania_3857_sf, ca_counties_sf)
```

## Summarize by county

Choropleth

```{r}
sesb_counts_sf <- ca_sesb_sf %>%
  group_by(county) %>%
  summarize(n_records = sum(!is.na(id)))

ggplot() +
  geom_sf(data = sesb_counts_sf, aes(fill = n_records), color = 'grey90', size = 1) +
  scale_fill_gradientn(colors = c('lightgrey', 'orange', 'purple')) +
  theme_minimal() +
  labs(fill = 'Number of S. punicea records')
  
```

## Next analysis:

For the county with the greatest number of red sesbania records, make a map of those locations and where they occur within the county.

### Pseudocode:

figure out which county has the most records
filter by just that county
take subset of other location information - watershed or location
plot the entire map or just plot the county with the counts, and add the subset of point locations
maybe use other spatial join dataset - sesb_ca_sf

```{r}
county_max <- sesb_counts_sf %>%
  filter(n_records == max(n_records)) %>%
  pull(county)

solano_sesb_sf <- sesb_ca_sf %>%
  filter(county %in% county_max)

solano_sf <- ca_counties_sf %>%
  filter(county %in% county_max)

ggplot() +
  geom_sf(data = solano_sf) +
  geom_sf(data = solano_sesb_sf, color = 'red')
```

# Making an interactive map with `tmap`

```{r}
#| eval: false
#| include: true

### set the viewing mode to interactive
tmap_mode(mode = 'view')

tm_shape(ca_counties_sf) +
  tm_fill("land_km2", palette = "BuGn") +
  tm_shape(sesbania_sf) +
  tm_dots()
  
```







